\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-11-14 paper/main.tex
added discussion of "tension" between different types of clustering

Probability of first recall curves~\citep{AtkiShif68, PostPhil65, WelcBurn24}
reflect the probability that an item will be recalled first, as a function of
its serial position during encoding. We used an analogous approach to compute
the proportion of trials on which each item (as a function of its presentation
position) was recalled at output position $n$~\citep{HowaKaha99, PolyEtal09,
ZhanEtal23}. To carry out this analysis, we initialized (for each participant)
a number-of-lists (16) by number-of-words-per-list (16) matrix of 0s. Then, for
each list, we found the index of the word that was recalled first, and we
filled in that position in the matrix with a 1. Finally, we averaged over the
rows of the matrix to obtain a 1 by 16 array of probabilities, for each
participant. We used an analogous procedure to compute probability of
$n$\textsuperscript{th} recall curves for each participant. Specifically, we
filled in the corresponding matrices according to the $n$\textsuperscript{th}
recall on each list that each participant made. When a given participant had
made fewer than $n$ recalls for a given list, we simply excluded that list from
our analysis when computing that participant's curve(s). The probability of
first recall curve corresponds to a special case where $n = 1$.

@@

>>>>

Probability of first recall curves~\citep{AtkiShif68, PostPhil65, WelcBurn24}
reflect the probability that an item will be recalled first, as a function of
its serial position during encoding. We used an analogous approach to compute
the proportion of trials on which each item (as a function of its presentation
position) was recalled at output position $n$~\citep{Hoga75, HowaKaha99,
PolyEtal09, ZhanEtal23}. To carry out this analysis, we initialized (for each
participant) a number-of-lists (16) by number-of-words-per-list (16) matrix of
0s. Then, for each list, we found the index of the word that was recalled
first, and we filled in that position in the matrix with a 1. Finally, we
averaged over the rows of the matrix to obtain a 1 by 16 array of
probabilities, for each participant. We used an analogous procedure to compute
probability of $n$\textsuperscript{th} recall curves for each participant.
Specifically, we filled in the corresponding matrices according to the
$n$\textsuperscript{th} recall on each list that each participant made. When a
given participant had made fewer than $n$ recalls for a given list, we simply
excluded that list from our analysis when computing that participant's
curve(s). The probability of first recall curve corresponds to a special case
where $n = 1$.

@@

\subsection*{How do different types of clustering relate to each other, and to memory performance?}

When the words on a studied list are presented in a random order, different types
of clustering in participants' recalls often tend to be negatively correlated.
For example, words that occur nearby on the list will not (on average)
tend to be semantically related, and vice versa. Therefore a participant who
shows a strong tendency to temporally cluster their recalls will tend to show weaker
semantic clustering, and so on~\citep{HealUitv19, HowaKaha02b, SedeEtal10}.  Further,
there is some evidence that temporal clustering is positively correlated with
memory performance, whereas semantic clustering is negatively correlated with
memory performance~\citep{SedeEtal10}.

The notion of ``multiple pathways to recall'' discussed above (see
\textit{Memory consequences of feature variability}) suggests one potential
explanation for these patterns. For example, temporal clustering has been
proposed to reflect reliance on contextual cues in an ``episodic'' pathway to
search memory, whereas semantic clustering reflects a relies on specific item
features. These two pathways may ``compete'' with each other during
recall~\citep{SochEtal09}. Meanwhile, extra-list intrusion errors (i.e., false
``recalls'' of items that were never encountered on the list) often tend to
share semantic features with recently recalled items~\citep{ZaroEtal06} and
also often lead the participant to stop recalling additional
items~\citep{MillEtal12a}. Speculatively, over-reliance on semantic cues may lead to
more intrusion errors, which in turn may lead to fewer recalls overall.

Our findings extend these prior results to consider lists that are \textit{not}
ordered randomly. Because ordering the words on a list along a particular
feature dimension removes the ``conflict'' between temporal and feature
clustering, the order manipulation conditions in our study represent an ``edge
case'' whereby different pathways to recall are not necessarily in conflict
with each other. For example, the same participants who exhibit strong feature
clustering \textit{also} show strong temporal clustering on ordered lists
(Fig.~\ref{fig:clustering-scatterplots}E). This is presumably at least partly
due to an inability to separate temporal and feature clustering on ordered
lists (also see \textit{Factoring out the effects of temporal clustering}).
However, features that change gradually with time (i.e., presentation position)
could also serve to strengthen the episodic (contextual) cues associated with
each item. In other words, participants might essentially combine multiple
noisy measures of change to form a more stable internal representation of
temporal context.
